### 2018-05-26 4_impl.tex
Cleanup finale up to Chap 3.

%TODO: ATTENTION HERE TO THE PICTURE!!!!!!!!!!!!!!!!!!!!!!!!
\begin{figure}[H]
  \centering
  %draft=true
  \includegraphics[height=.83\textheight,keepaspectratio]{img/my/draw.io/general_arch-no_numbering.png} %{img/my/draw.io/general_arch.png}
  \caption{The general architecture of \porthos[2]}
  \label{fig:arch}
\end{figure}

\subsection{Program input}
\label{ch:impl:input}
%(for instance, the original input language of \porthos{} tool, two variants of syntax of a litmus test used by \tool{herd}, an assembly language for any supported architecture).

@@

@@

The \ytree{} is compiled into the low-level event-based program representation called \textit{\xgraph{}}.
The mathematical structure of event-flow graph was discussed in Section~\ref{ch:wmm:event}.
%In general, \xgraph{} follows the structure described there, however it is constructed in several stages.
The nodes of the graph are events, and the edges are basic relations: the control-flow relation \po{} and the data-flow relations \co{} and \rf{}.
Hereinafter, we denote the \xgraph{} with only control-flow edges as \xgraph[CF], the \xgraph{} with only data-flow edges as \xgraph[CF].
The complete \xgraph{} is $\xgraph[CF+DF] = \xgraph[CF] \; \cup \; \xgraph[DF]$.
The UML diagram on Figure~\ref{fig:class-diagrams:XEntity-interfaces} represents the hierarchy of main interfaces of the X-abstraction level.
Internally, the graph is represented by an adjacency matrix (to be exact, by multiple adjacency matrices that store edges of different kinds, see more details in Section~\ref{ch:impl:model:xgraph:edges}).

@@

The \xgraph{} abstract machine model disposes infinitely many memory units, both local and shared (see Section~\ref{ch:impl:model:xgraph:mem}).
The X-compiler accesses all memory units via the \texttt{XMemoryManager}, which by the end of pre-compilation stage is already initialised (has registered all shared memory units).
However, the memory manager is a stateful component of the compiler as it offers the methods for declaring and removing local memory units dynamically at the compilation time.

@@

\begin{figure}[!h]
\begin{lstlisting}
class XKernelInvocationHook
       extends XInvocationHookBase 
       implements XInvocationHook {
 ...
 @Override
 public XInvocationHookAction intercept(String functionName) {
   switch (functionName) {
     case "WRITE_ONCE": {
       return new XInvocationHookAction((receiver, arguments) -> {
         if (arguments.length != 2 || receiver != null) {
             return null;  // do not intercept
         }
         XMemoryUnit destUnit = arguments[0];
         if (!(destUnit instanceof XSharedLvalueMemoryUnit)) {
           throw new XMethodInvocationError(methodName, 
               "arg 1: not a shared l-value memory unit: " + destUnit);
         }
         XSharedLvalueMemoryUnit dest = (XSharedLvalueMemoryUnit) destUnit;
         XMemoryUnit srcUnit = arguments[1];
         if (!(srcUnit instanceof XLocalMemoryUnit)) {
           throw new XMethodInvocationError(methodName, 
               "arg 2: not a local memory unit: " + srcUnit);
         }
         XLocalMemoryUnit src = (XLocalMemoryUnit) srcUnit;
         return program.emitMemoryEvent(dest, src);
     });
   }
   case "READ_ONCE": {
     ...
   }
 }
}
\end{lstlisting}
\caption{Example of the invocation hook for intercepting the Linux kernel-specific functions}
\label{code-invoc-hook}
\end{figure}

@@

%In this paragraph we discuss only the implementation of the process interpreter \texttt{XProcessInterpreter}.
%This interpreter is used to build the branched event-flow graph with all kinds of jumps, memory operations and fences.
The X-program interpreter is invoked by the \texttt{Y2XConverterVisitor} which walks down the recursive syntax tree \ytree{}.
The calls to the X-program interpreter are dispatched to currently maintained X-process interpreter.
To be able to properly recognise nested statements of the \ytree{}, the X-process interpreter needs to have stacks.
On any semantic error, the X-interpreter throws an \texttt{XInterpretationError}.

@@

All interface methods can be divided into two groups.
The first group constitute methods that emit events.
These methods construct an event object (as events contain the \texttt{XEventInfo} structure that stores information about the owning process, all events must be created by the process constructor, i.e. X-interpreter) and change the state of interpreter considering the newly created event.
Note that the methods \texttt{createComputationEvent} do not emit a computation event but only create one.
This is performed as optimisation that removes unused computations from the model (otherwise, for instance, the assignment 
`\texttt{x = 1 * (2 + 3)}' would be compiled to two consequent events `\texttt{eval(2 + 3); write(register <- eval(1 * eval(2 + 3)));}' instead of a single event `\texttt{write(register <- eval(1 * eval(2 + 3)));}').
If the computation event was not used at all (for example, in the following C code: `\texttt{foo(); 1; bar();}' the execution event `\texttt{eval(1)}' is skipped by the model), it is also removed from the event-graph (see justification in Section~\ref{ch:impl:model:xgraph:evt}).
The second group of X-interpreter methods consists of the methods for defining non-linear statements (branchings and loops).
These methods change the state of the interpreter and set up additional non-linear control-flow edges.
%`\texttt{register = 1 * (location + 2)}' would be compiled to three consequent events `\texttt{load(temp\_reg\_1 <- location); eval(temp\_reg\_1 + 2); write(register <- 1 * (temp\_reg\_1 + 2));}',

As a high-level instruction (Y-level) may be compiled into a sequence of low-level instructions (for example, a computation that involves shared variables should firstly load them into the local memory and then process the computation event over local-only memory), the interpreter must maintain the \textit{stack of contexts} and remember the \textit{previous event} to be able to correctly process nested non-linear statements of C language.
The context is a data structure that carries the \textit{state} and some additional information in the case of processing non-linear statements (e.g., conditional event, first and last then- and else-branch events for binding, etc.).
Once the new event has been emitted, the interpreter sets up control-flow edges w.r.t. state of the context on the top of the stack (the context stack is always non empty: the bottom context is a linear one).
The context state is an enumeration of the following values:
\begin{itemize}
\item \texttt{WaitingAdditionalCommand}: the interpreter is in the state of defining the complex (non-linear) statement and is not able to process any new event (will throw an exception if any) until the state is not changed;
\item \texttt{WaitingFirstConditionEvent}: the first next emitted event will be accepted as the first event of the condition evaluation (later, the loop edges will be set to this event);
\item \texttt{WaitingLastConditionEvent}: the next event must be of type \texttt{XComputationEvent}; it will be saved as the conditional branching event;
\item \texttt{WaitingFirstSubBlockEvent}: the first next emitted event will be accepted as the first event of the branching statement (later the interpreter will set up jumps to that event);
\item \texttt{WaitingNextLinearEvent}: the standard interpreter state for processing next linear event, and
\item \texttt{Idle}: the interpreter does not set up the edge from the previous event to the new event.
\end{itemize}

@@

@@

For the sake of simplicity of the interpreter, jump events (no-computation events) are present in the flow-graph, however they can be removed from it with rebinding ingoing and outgoing edges.

@@

In order to be encoded into an SMT-formula, the compiled graph needs to be acyclic~\cite{Porthos17b}.
To convert it to an acyclic form, we perform the \textit{unrolling} (sometimes called \textit{unwinding}) transfomation: each cycle is unrolled up to the user-defined bound $k$.
Comparing to \porthos[1], we have changed the meaning of the bound: instead of executing all cycles at most $k$ times, the \porthos[2] interprets the bound as the maximum number of events in the trace of the unrolled graph%
%
\footnote{The original specification of \porthos[2] stated that the unrolling bound $k$ must be interpreted as the maximum number of instruction in the original code (technically, the number of expressions in the ANTLR syntax tree).
Nonetheless, current implementation of the \xgraph{} unroller counts the X-level events as it is much simpler to implement (the questions about how to count complex expressions that involve multiple shared variables and method invocations are left open for the future versions of \porthos[2]).
For now, the information about the element of the ANTLR tree that corresponds to the X-event or Y-tree element is being lost during the transformations.
Instead, we use the \texttt{LocationService} discussed in Section~\ref{ch:impl:model:ytree} that in perspective may be extended for providing this information along with the text citation of the original code.}
%
.
The meaning of a bound was changed in order to increase predictability of the size of the result SMT-formula.
Not that recursive function calls create unconditional jumps back in the event-flow graph, which will also be recognised as a loop and be affected by the unrolling algorithm.

@@

Once the graph is unrolled, it can be augmented by data-flow edges.
As it was discussed in Paragraph~\ref{ch:impl:model:xgraph:edges}, the data-flow edges can represent either \rf{}- or \co{}-relation.
The \rf{}-edges join each write event to all read events that access the same location; the \co{}-edges join write events to the same location.

>>>>

\section{Program input}
\label{ch:impl:input}
%(for instance, the original input language of \porthos{} tool, two variants of syntax of a litmus test used by \tool{herd}, an assembly language for any supported architecture).

@@

%TODO: ATTENTION HERE TO THE PICTURE!!!!!!!!!!!!!!!!!!!!!!!!
\begin{figure}[H]
  \centering
  \includegraphics[height=.95\textheight,keepaspectratio]{img/my/draw.io/general_arch-no_numbering.png}
  \caption{The general architecture of \porthos[2]}
  \label{fig:arch}
\end{figure}

@@

The \ytree{} is compiled into the low-level event-based program representation called \textit{\xgraph{}}.
The mathematical structure of event-flow graph was discussed in Section~\ref{ch:wmm:event}.
%In general, \xgraph{} follows the structure described there, however it is constructed in several stages.
The nodes of the graph are events, and the edges are basic relations: the control-flow relation \po{} and the data-flow relations \co{} and \rf{}.
Hereinafter, we denote the \xgraph{} with only control-flow edges as \xgraph[CF], the \xgraph{} with only data-flow edges as \xgraph[CF].
The complete \xgraph{} is $\xgraph[CF+DF] = \xgraph[CF] \; \cup \; \xgraph[DF]$.
The UML diagram on Figure~\ref{fig:class-diagrams:XEntity-interfaces} represents the hierarchy of main interfaces of the X-abstraction level.
Internally, the graph is represented by an adjacency matrix (to be exact, by multiple adjacency matrices that store edges of different kinds, see more details in Section~\ref{ch:impl:model:xgraph}).

@@

The \xgraph{} abstract machine model disposes infinitely many memory units, both local and shared (see Section~\ref{ch:impl:model:xgraph}).
The X-compiler accesses all memory units via the \texttt{XMemoryManager}, which by the end of pre-compilation stage is already initialised (has registered all shared memory units).
However, the memory manager is a stateful component of the compiler as it offers the methods for declaring and removing local memory units dynamically at the compilation time.

@@

\begin{figure}[!h]
\begin{lstlisting}
class XKernelInvocationHook extends XInvocationHookBase 
       implements XInvocationHook {
 ...
 @Override
 public XInvocationHookAction intercept(String functionName) {
   switch (functionName) {
     case "WRITE_ONCE": {
       return new XInvocationHookAction((receiver, arguments) -> {
         if (arguments.length != 2 || receiver != null) {
           return null;  // do not intercept
         }

         XMemoryUnit destUnit = arguments[0];
         if (!(destUnit instanceof XSharedLvalueMemoryUnit)) {
           throw new XMethodInvocationError(methodName, 
               "arg 1: not a shared l-value memory unit: " + destUnit);
         }
         XSharedLvalueMemoryUnit dest = 
               (XSharedLvalueMemoryUnit) destUnit;

         XMemoryUnit srcUnit = arguments[1];
         if (!(srcUnit instanceof XLocalMemoryUnit)) {
           throw new XMethodInvocationError(methodName, 
               "arg 2: not a local memory unit: " + srcUnit);
         }
         XLocalMemoryUnit src = (XLocalMemoryUnit) srcUnit;

         return program.emitMemoryEvent(dest, src);
     });
   }
   case "READ_ONCE": {
     ...
   }
 }
}
\end{lstlisting}
\caption{Example of the invocation hook for intercepting the Linux kernel-specific functions}
\label{code-invoc-hook}
\end{figure}

@@

%TODO: say that program is the set of processes , but the implements the same interface !!! it's not clear here.

%TODO: !!! Say that interpreter does not know anyth about Y, all its interface methods are <sth> -> <X>. The only part that know about Y is Y2Xvisitor

%In this paragraph we discuss only the implementation of the process interpreter \texttt{XProcessInterpreter}.
%This interpreter is used to build the branched event-flow graph with all kinds of jumps, memory operations and fences.
The X-program interpreter is invoked by the \texttt{Y2XConverterVisitor} which walks down the \ytree{} recursively.
The calls to the X-program interpreter are dispatched to currently maintained X-process interpreter.
To be able to properly recognise nested statements of the \ytree{}, the X-process interpreter needs to have stacks.
On any semantic error, the X-interpreter throws an \texttt{XInterpretationError}.

@@

All interface methods can be divided into two groups.
The first group constitute methods that emit events.
These methods construct an event object (as events contain the \texttt{XEventInfo} structure that stores information about the owning process, all events must be created by the process constructor, i.e. X-interpreter) and change the state of interpreter considering the newly created event.
Note that the methods \texttt{createComputationEvent} do not emit a computation event but only create one.
This is performed as optimisation that removes unused computations from the model (otherwise, for instance, the assignment 
`\texttt{x = 1 * (2 + 3)}' would be compiled to two consequent events `\texttt{eval(2 + 3); write(register <- eval(1 * eval(2 + 3)));}' instead of a single event `\texttt{write(register <- eval(1 * eval(2 + 3)));}').
If the computation event was not used at all (for example, in the following C code: `\texttt{foo(); 1; bar();}' the execution event `\texttt{eval(1)}' is skipped by the model), it is also removed from the event-graph (see justification in Section~\ref{ch:impl:model:xgraph}).
The second group of X-interpreter methods consists of the methods for defining non-linear statements (branchings and loops).
These methods change the state of the interpreter and set up additional non-linear control-flow edges.
%`\texttt{register = 1 * (location + 2)}' would be compiled to three consequent events `\texttt{load(temp\_reg\_1 <- location); eval(temp\_reg\_1 + 2); write(register <- 1 * (temp\_reg\_1 + 2));}',

As a high-level (Y-level) instruction may be compiled into a sequence of low-level instructions (for example, a computation that involves shared variables should firstly load them into the local memory and then process the computation event over local-only memory), the interpreter must maintain the \textit{stack of contexts} and remember the \textit{previous event} to be able to correctly process nested non-linear statements of C language.
The context is a data structure that carries the \textit{state} and some additional information in the case of processing non-linear statements (e.g., conditional event, first and last then- and else-branch events for binding, etc.).
Once the new event has been emitted, the interpreter sets up control-flow edges w.r.t. state of the context on the top of the stack (the context stack is always non empty: the bottom context is a linear one).
The context state is an enumeration of the following values:

\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{WaitingAdditionalCommand}: the interpreter is in the state of defining the complex (non-linear) statement and is not able to process any new event (will throw an exception if any) until the state is not changed;
\item \texttt{WaitingFirstConditionEvent}: the first next emitted event will be accepted as the first event of the condition evaluation (later, the loop edges will be set to this event);
\item \texttt{WaitingLastConditionEvent}: the next event must be of type \texttt{XComputationEvent}; it will be saved as the conditional branching event;
\item \texttt{WaitingFirstSubBlockEvent}: the first next emitted event will be accepted as the first event of the branching statement (later the interpreter will set up jumps to that event);
\item \texttt{WaitingNextLinearEvent}: the default interpreter state for processing next linear event, and
\item \texttt{Idle}: the interpreter does not set up the edge from the previous event to the new event.
\end{itemize}

@@

As an example of usage the interpreter, consider the code of the \texttt{Y2XConverterVisitor} for processing \texttt{if-then-else} statement of the \ytree{} in Figure~\ref{TODO}.

\begin{figure}
%TODO!!! refactor the code, move program.tryEvaluateComputation() to Y2Xvisitor or Helper
\begin{lstlisting}
public class Ytree2XgraphConverterVisitor implements YtreeVisitor<XEntity> {
  ...

  @Override
  public XEvent visit(YBranchingStatement node) {
    program.startBlockDefinition(XInterpreter.BlockKind.Branching);

    program.startBlockConditionDefinition();
    XComputationEvent condition = program.tryEvaluateComputation(node.getCondition().accept(this));
    program.finishBlockConditionDefinition(condition);

    program.startBlockBranchDefinition(XInterpreter.BranchKind.Then);
    node.getThenBranch().accept(this);
    program.finishBlockBranchDefinition();

    YStatement elseBranch = node.getElseBranch();
    program.startBlockBranchDefinition(XInterpreter.BranchKind.Else);
    if (elseBranch != null) {
      elseBranch.accept(this);
    }
    else {
      program.emitNopEvent(); // needed for encoding
    }
    program.finishBlockBranchDefinition();

    program.finishNonlinearBlockDefinition();

    return null; //statements return null
    }
}
\end{lstlisting}
\end{figure}

@@

For the sake of simplicity of the interpretation and encoding, jump events (no-computation events) are present in the flow-graph, however they can be removed from it with rebinding ingoing and outgoing edges.

@@

In order to be encoded into an SMT-formula, the compiled graph needs to be acyclic~\cite{Porthos17b}.
To convert it to an acyclic form, we perform the \textit{unrolling} (sometimes called \textit{unwinding}) transfomation: each cycle is unrolled up to the user-defined bound $k$.
Comparing to \porthos[1], we have changed the meaning of the bound: instead of executing all cycles at most $k$ times, the \porthos[2] interprets the bound as the maximum number of events in the trace of the unrolled graph%
%
\footnote{The original specification of \porthos[2] stated that the unrolling bound $k$ must be interpreted as the maximum number of instruction in the original code (technically, the number of expressions in the ANTLR syntax tree).
Nonetheless, current implementation of the \xgraph{} unroller counts the X-level events as it is much simpler to implement (the questions about how to count complex expressions that involve multiple shared variables and method invocations are left open for the future versions of \porthos[2]).
For now, the information about the element of the ANTLR tree that corresponds to the X-event or Y-tree element is being lost during the transformations.
Instead, we use the \texttt{LocationService} discussed in Section~\ref{ch:impl:model:ytree} that in perspective may be extended for providing this information along with the text citation of the original code.}
%
.
The meaning of a bound was changed in order to increase predictability of the size of the result SMT-formula.
Note that recursive function calls create unconditional back-jumps in the event-flow graph, which will also be recognised as a loop and be affected by the unrolling algorithm.
The discussion on how to set up the \po-relations in the event-flow graph with recursive function calls can be found in~\cite{musketeer2014recursion}.

@@

Once the graph is unrolled, it can be augmented by data-flow edges.
As it was discussed in Section~\ref{ch:impl:model:xgraph}, the data-flow edges can represent either \rf{}- or \co{}-relation.
The \rf{}-edges join each write event to all read events that access the same location; the \co{}-edges join write events to the same location.
